---
title: "Using ggAtlasTrack"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using ggAtlasTrack}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 4,
  fig.align = "center"
)
```

# Overview

`ggAtlasTrack` provides `ggplot2`-based visualisation tools for AtlasTrack white-matter tract polygons, inspired by the `ggseg` ecosystem (Mowinckel & Vidal-Piñeiro, 2020). `ggAtlasTrack` follows a similar plotting philosophy for AtlasTrack white-matter tract visualisation, while providing package-specific view sets and tract-level data mapping. The package can be used to visualise the AtlasTrack tract atlas or to map tract-level continuous values, such as coefficients, association statistics, or significance scores.

```{r setup}
library(ggAtlasTrack)
```

# Available view sets

The available plotting views can be listed using `atlastrack_views_available()`.

```{r available-views}
atlastrack_views_available()
```

# Plotting the default view

By default, `atlastrack_plot()` displays the orthogonal view set.

```{r orthogonal-plot}
atlastrack_plot()
```

This is equivalent to:

```{r orthogonal-code-only, eval = FALSE}
atlastrack_plot(view_set = "orthogonal")
```

# Plotting the ggseg-style view

The `"ggseg"` view displays a split axial/coronal layout inspired by the visual style of `ggseg`.

```{r ggseg-plot}
atlastrack_plot(view_set = "ggseg")
```

# Plotting the axial view

The `"axial"` view displays inferior, middle, and superior axial slices. This can be useful for visualising inferior/anterior tracts, such as the uncinate fasciculus, which cannot be seen in the default orthogonal or ggseg views.

```{r axial-plot}
atlastrack_plot(view_set = "axial")
```

# Listing available tract names

The available tract abbreviations and full tract names can be listed using `atlastrack_tract_names()`.

```{r tract-names}
atlastrack_tract_names()
```

# Plotting tract-level values

Users can provide a data frame containing one row per tract and a continuous value to map onto the atlas.

In this example, we create random tract-level values called `feature_importance`.

```{r example-data}
set.seed(1)

tract_values <- tibble::tibble(
  tract_name = unique(atlastrack_tracts$tract_name),
  feature_importance = runif(
    length(unique(atlastrack_tracts$tract_name)),
    0,
    1
  )
)

head(tract_values)
```

The `value` argument specifies which column should be mapped to tract fill.

```{r feature-importance-plot}
atlastrack_plot(
  data = tract_values,
  value = feature_importance,
  view_set = "orthogonal",
  legend_title = "Feature\nimportance"
)
```

# Setting colour scale limits

By default, the colour scale is based on the range of the supplied data. To anchor the scale at zero, use the `limits` argument.

```{r feature-importance-plot-limits}
atlastrack_plot(
  data = tract_values,
  value = feature_importance,
  view_set = "orthogonal",
  limits = c(0, max(tract_values$feature_importance, na.rm = TRUE)),
  legend_title = "Feature\nimportance"
)
```

# Customising colours

The low and high colours can be changed using the `low` and `high` arguments.

```{r custom-colours}
atlastrack_plot(
  data = tract_values,
  value = feature_importance,
  view_set = "axial",
  low = "grey95",
  high = "#2166AC",
  limits = c(0, max(tract_values$feature_importance, na.rm = TRUE)),
  legend_title = "Feature\nimportance"
)
```

# Hiding ventricles

Ventricles are displayed by default. They can be hidden using `show_ventricles = FALSE`.

```{r hide-ventricles}
atlastrack_plot(
  view_set = "orthogonal",
  show_ventricles = FALSE
)
```

# Saving figures

Because `atlastrack_plot()` returns a `ggplot` object, figures can be saved using `ggplot2::ggsave()`.

```{r save-figure, eval = FALSE}
p <- atlastrack_plot(view_set = "axial")

ggplot2::ggsave(
  filename = "atlastrack_axial.png",
  plot = p,
  width = 8,
  height = 4,
  dpi = 300
)
```

# Notes on interpretation

These plots are two-dimensional visualisations of preprocessed AtlasTrack tract polygons. The view sets are designed to balance anatomical readability and tract coverage.

Some tracts may be more clearly visible in one view set than another. For example, inferior/anterior tracts may be better visualised in the axial view than in the default orthogonal view.

# References

Hagler, D. J., Jr., Ahmadi, M. E., Kuperman, J., Holland, D., McDonald, C. R., Halgren, E., & Dale, A. M. (2009). Automated white-matter tractography using a probabilistic diffusion tensor atlas: Application to temporal lobe epilepsy. *Human Brain Mapping, 30*(5), 1535–1547. https://doi.org/10.1002/hbm.20619

Mowinckel, A. M., & Vidal-Piñeiro, D. (2020). Visualization of brain statistics with R packages ggseg and ggseg3d. *Advances in Methods and Practices in Psychological Science, 3*(4), 466–483. https://doi.org/10.1177/2515245920928009



